---
title: "Data types and structures"
subtitle: "ACTEX Learning - AFDP: R Session"
format: 
  html:
    toc: true
editor: visual
execute:
  warning: false
  message: false
---

```{r}
#| echo: false
library(ggplot2)
book_theme <- theme_minimal() + 
  theme(plot.title=element_text(face="bold"))
ggplot2::theme_set(book_theme)
```

## Getting into Practice with R

`Data manipulation` is a fundamental skill in the field of `data science` and the `actuarial profession`. It involves the process of `transforming`, `organizing`, and `cleaning` raw data to make it suitable for tailored analysis and visualization. 

In the actuarial context, tools like `R` are invaluable for tasks such as: calculating `premium rates` for mortality and morbidity products, `evaluating the probability` of financial loss or return, providing business `risk consulting`, and `planning` for pensions and retirement. In essence, `R` is a powerful tool for actuaries and data scientists to perform complex data analysis and modeling.

## R Packages

R is equipped with a vast ecosystem of packages that extend its functionality.

R packages are collections of functions, data, and compiled code in a well-defined format. They are stored in a directory called `library`. A fresh R installation includes a set of packages that are loaded automatically when R starts. These packages are referred to as the `base` packages. The base packages are always available, and they do not need to be loaded explicitly.

Others packages are available for download from [CRAN (Comprehensive R Archive Network)](https://cran.r-project.org/web/packages/available_packages_by_name.html) or other repositories, such as [GitHub](https://github.com/) usually for package development versions. To install a package which is stored on CRAN, we can use the `install.packages("")` function. To load a package, use the `library()` function.

## Data Types

R has several data types, including numeric, character, logical, integer, and complex. The most common data types are:

-   `Numeric` - real numbers
-   `Character` - text
-   `Logical` - TRUE or FALSE
-   `Integer` - whole numbers

It allows data to be on the form of:

-   `Vectors` of numbers, characters, or logical values
-   `Matrices` which are arrays with two dimensions
-   `Arrays` which are multi-dimensional generalizations of matrices
-   `Data Frames` which are matrices with columns of different types
-   `Lists` which are collections of objects

## Basic R Syntax

R is a case-sensitive language, meaning that it distinguishes between uppercase and lowercase letters. It uses the `#` symbol to add comments to the code. Comments are ignored by the R interpreter and are used to explain the code.

```{r}
# Basic arithmetic and variable assignment
x <- 10
y <- 5
sum_xy <- x + y
sum_xy
```

```{r}
# Simple interest formula
P <- 1000  # Principal
r <- 0.05  # Interest rate
t <- 2     # Time in years
A <- P * (1 + r * t)
A
```

## R Functions

R is a functional programming language, which means that it is based on functions. Functions are blocks of code that perform a specific task. They take input, process it, and return output.

There are two types of functions in R: `base functions` and `user-defined functions`. Base functions are built into R, while user-defined functions are created by the user.

### Base Functions

For example, the `mean()` function calculates the average of a set of numbers. The `mean()` function takes a vector of numbers as input and returns the average of those numbers. To access a general function information in R, use the `help()` function, or the `?` operator.

```{r}
#| eval: false
?mean()
help(mean)
```

The `c()` function is used to combine values into a vector or list.

```{r}
mean(c(1, 2, 3, 4, 5))
```

Useful base R function:

-   `mean()` - calculates the average of a set of numbers
-   `sum()` - calculates the sum of a set of numbers
-   `sd()` - calculates the standard deviation of a set of numbers
-   `var()` - calculates the variance of a set of numbers
-   `min()` - returns the minimum value in a set of numbers
-   `max()` - returns the maximum value in a set of numbers
-   `length()` - returns the length of a vector
-   `str()` - displays the structure of an R object
-   `class()` - returns the class of an object
-   `typeof()` - returns the type of an object
-   `summary()` - provides a summary of an object
-   `plot()` - creates a plot of data ...

### User-Defined Functions

```{r}
#| eval: false
name <- function(variables) {
  # Code block
  }
```

```{r}
x <- c(1, 2, 3, 4, 5)

avg <- function(x) {
  sum(x)/length(x)
  }
```

```{r}
avg(x);
mean(x)
```

## Essential Data Preparation Techniques

Data wrangling, manipulation, and transformation are essential techniques for preparing and refining data for analysis. A series of steps to ensure data quality by handling inconsistencies, filling gaps, correcting errors, removing duplicates, and merging dataset. Common data preparation tasks include:

- Wrangling: 

  - Handling missing values
  - Reshaping data (wide ↔ long)
  - Removing duplicates
  - Standardizing formats (e.g., dates, units)
  - Joining multiple datasets

- Manipulation:

  - Selecting and filtering rows/columns (select(), filter())
  - Sorting (arrange())
  - Grouping and summarizing (group_by(), summarise())

- Transformation:

  - Normalization or scaling (e.g., z-scores, min-max)
  - Encoding categorical variables
  - Creating new variables (mutate())
  - Applying functions (e.g., log, square root)

These processes, often performed using R packages like {dplyr}, {tidyr}, and {data.table}, are crucial steps in converting raw, disorganized data into meaningful insights. Together, they form the foundation for effective data analytics, which uses cleaned and well-structured data to perform descriptive, predictive, and inferential analyses.

As an example, consider a dataset with columns for `ID`, `Age`, and `Salary`. We can remove rows with missing values in the `Age` column and create a new column called `IncomeGroup` based on the `Salary` column.

```{r}
# Create a data frame
data <- data.frame(
  ID = 1:5,
  Age = c(25, 30, NA, 45, 35),
  Salary = c(50000, 60000, 55000, NA, 70000)
)

data
```

Handling missing values and creating a new variable can be done using the `tidyverse` set of packages. The `dplyr` package provides functions (or verbs) for data manipulation, such as `filter()`, `mutate`, `select()`, and more. The `magrittr` package provides the pipe operator (`%>%`) used to chain functions together, making the code more readable. Furthermore, the pipe operator has recently been included in base R as well, so called the "tee" operator (`|>`) or native pipe operator.

Load the `tidyverse` package:

```{r}
library(tidyverse)
```

```{r}
cleaned_data <- data %>%
  # Remove rows with missing Age values
  filter(!is.na(Age)) %>%   
  # Create a new column based on Salary
  mutate(IncomeGroup = ifelse(Salary > 60000, "High", "Low")) 

cleaned_data
```

An interesting function to use is from the `janitor` package. The `clean_names()` function cleans the column names of a data frame by converting them to lowercase, replacing spaces with underscores, and removing special characters.

```{r}
cleaned_data %>%
  # Clean column names (here the janitor package is called with ::)
  janitor::clean_names() 
```

R is a powerful tool for actuaries and data scientists, it offers a wide range of functions and packages for data analysis, modeling, and visualization. By mastering R programming, actuaries can enhance their analytical skills, improve decision-making, and drive innovation in the insurance industry.


## Resources

-   Charpentier, A., & Kaas, R. (2014). [Computational Actuarial Science with R. CRC Press](https://www.amazon.it/Computational-Actuarial-Science-Arthur-Charpentier/dp/1466592591)
-   Wickham, H. (2016). [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/)
-   Tidyverse (2021). [Tidyverse: Easily Install and Load the 'Tidyverse'](https://www.tidyverse.org/)
